---
title: "Lab02: Distances and the Border Zone"
subtitle: 'Ecosystem Science and Sustanability 523c'
author: 
  name: Billy Johnson
  email: billyjohsnon181@gmail.com
format: html
---

# Load in the libraries
```{r, message=FALSE}
# spatial data science
library(tidyverse)
library(sf)
library(units)

# Data
library(USA.state.boundaries)
library(rnaturalearth)

# Visualization
library(gghighlight)
library(ggrepel)
library(knitr)
library(flextable)
library(leaflet)
library(ggthemes)

# Other
library(readr)
```

# Question 1

### 1.1 Define a projection
Use North America Equidistant Conic
```{r}
eqdc <- '+proj=eqdc +lat_0=40 +lon_0=-96 +lat_1=20 +lat_2=60 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs'
```

### 1.2 Get USA state boudaries
```{r}
remotes::install_github("ropensci/USAboundaries")
remotes::install_github("ropensci/USAboundariesData")

# Once installed
USA_states_raw <- USAboundaries::us_states(resolution = "low")
```

### 1.3 Get country boundaries for Mexico, the US and Canada
```{r}
remotes::install_github("ropenscilabs/rnaturalearthdata")

countries <- rnaturalearth::countries110 %>% 
  st_as_sf() %>% 
  filter(countries110$ADMIN %in% c("United States of America", "Canada", "Mexico")) %>% 
  st_transform(crs = eqdc)
```

### 1.4 Get city locations from the csv file
```{r}
city_locations <- read_csv("Lab2/simplemaps_uscities_basicv1/uscities.csv")

city_locations_clean <- city_locations %>% 
  filter(!state_id %in% c("AK", "HI", "PR"))

# Convert to spatial
city_location_sp <- st_as_sf(city_locations_clean,
                             coords = c("lng", "lat"),
                             crs = 4326) %>% 
  select(city, population, state_name) %>% 
  st_transform(crs = eqdc)


#st_filter(city_location_sp,
#          filter(city_location_sp, city == "Fort Collins"),
#          .predicate = st_is_within_distance, 1000)
```

# Question 2
### 2.1 Distance to USA border (coastline or national) (km)
```{r}
# Convert USA state boundaries to a MULTILINESTRING
USA_border <- USA_states_raw %>% 
  filter(!state_abbr %in% c("AK", "HI", "PR")) %>% 
  st_union() %>% 
  st_cast("MULTILINESTRING") %>% 
  st_transform(crs = eqdc)

# CREATE DISTANCE COLUMN
city_location_sp$dist_us_border_km <- st_distance(city_location_sp, USA_border) %>% 
  set_units("km") %>% 
  drop_units()

# Create flextable
top5_us_border <- city_location_sp %>% 
  slice_max(order_by = dist_us_border_km, n = 5) %>% 
  select(city, state_name, dist_us_border_km) %>% 
  flextable() %>% 
  set_caption("Top 5 US cities with the greatest distance to the US border")

top5_us_border
```

### 2.2 Distance to state borders
```{r}
# Create US state borders
state_borders <- USA_states_raw %>% 
  filter(!state_abbr %in% c("AK", "HI", "PR")) %>% 
  st_combine() %>% 
  st_cast("MULTILINESTRING") %>% 
  st_transform(crs = eqdc) 

# Calculate the distances to state border
city_location_sp$dist_state_border_km <- st_distance(city_location_sp, state_borders) %>% 
  set_units("km") %>% 
  drop_units()

# Create table
top5_state_border <- city_location_sp %>% 
  slice_max(order_by = dist_state_border_km, n = 5) %>% 
  select(city, state_name, dist_state_border_km) %>% 
  flextable() %>% 
  set_caption("Top 5 US cities with the greatest distance to the state border")

top5_state_border
```

### 2.3 Distance to Mexico
```{r}
mexico <- countries %>% 
  filter(ADMIN == "Mexico") %>% 
  st_union() %>% 
  st_cast("MULTILINESTRING") %>% 
  st_transform(crs = eqdc)

city_location_sp$dist_mexico_km <- st_distance(city_location_sp, mexico) %>% 
  set_units("km") %>% 
  drop_units()

top5_mexico <- city_location_sp %>% 
  slice_max(order_by = dist_mexico_km, n = 5) %>% 
  select(city, state_name, dist_mexico_km) %>% 
  flextable() %>% 
  set_caption("Top 5 US cities longest distance to Mexico border")

top5_mexico
```

### 2.4 Distance to Canada (km)
```{r}
canada <- countries %>% 
  filter(ADMIN == "Canada") %>% 
  st_union() %>% 
  st_cast("MULTILINESTRING") %>% 
  st_transform(crs = eqdc)

city_location_sp$dist_canada_km <- st_distance(city_location_sp, canada) %>% 
  set_units("km") %>% 
  drop_units()

top5_canada <- city_location_sp %>% 
  slice_max(order_by = dist_canada_km, n = 5) %>% 
  select(city, state_name, dist_canada_km) %>% 
  flextable() %>% 
  set_caption("Top 5 US cities with the longest distance to the Canadian border")

top5_canada
```


# Question 3
Visualization of distance data
### 3.1 Data
```{r}
# Show the 3 continents, CONUS outline, state boundaries, and 10 largest USA cities (population) on a single map

top10_cities <- city_location_sp %>% 
  arrange(desc(population)) %>% 
  slice(1:10)

# Plot
ggplot()+
  geom_sf(data = countries, fill = "grey", color = "white", lty = "solid", size = 0.3)+
  geom_sf(data = USA_border, fill = NA, color = "black", lty = "dashed", size = 0.3)+
  geom_sf(data = state_borders, fill = NA, color = "black", lty = "solid", size = 0.05)+
  geom_sf(data = top10_cities, color = "red")+
  ggrepel::geom_label_repel(data = top10_cities, aes(label = city, geometry = geometry),
                            stat = "sf_coordinates",
                            size = 3)+
  theme_map()+
  labs(title = "Map of US top 10 most populated cities")

```

### 3.2 City Distance from the border
```{r}
top5_city_distance <- city_location_sp %>% 
  arrange(desc(dist_us_border_km)) %>%
  slice(1:5) %>% 
  mutate(city_label = paste0(city, ",", state_name))


ggplot()+
  geom_sf(data = city_location_sp, aes(color = dist_us_border_km))+
  scale_color_viridis_c(option = "plasma")+
  labs(color = "Distance to US Border (km)")+
  ggrepel::geom_label_repel(data = top5_city_distance, aes(label = city_label, geometry = geometry),
                            stat = "sf_coordinates",
                            size = 3)+
  theme_map()+
  labs(title = "Cities in the US and the distance to the U.S. Border")
```

### 3.3 City Distance from Nearest state
```{r}
top5_city_distance_state <- city_location_sp %>% 
  arrange(desc(dist_state_border_km)) %>%
  slice(1:5) %>% 
  mutate(city_label = paste0(city, ",", state_name))


ggplot()+
  geom_sf(data = city_location_sp, aes(color = dist_state_border_km))+
  scale_color_viridis_c(option = "plasma")+
  labs(color = "Distance to State Border (km)")+
  ggrepel::geom_label_repel(data = top5_city_distance_state, aes(label = city_label, geometry = geometry),
                            stat = "sf_coordinates",
                            size = 3)+
  theme_map()+
  labs(title = "City Distance to State Boundary")
```

### 3.4 Equidistance boundary from Mexico and Canada
```{r}
city_location_sp <- city_location_sp %>% 
  mutate(absolute_distance = abs(dist_mexico_km - dist_canada_km))

equal_distance_cities <- city_location_sp %>% 
  filter(absolute_distance <= 100)

top5_pop_cities_near_border <- equal_distance_cities %>% 
  arrange(desc(population)) %>% 
  slice_head(n =5)

ggplot()+
  geom_sf(data = city_location_sp, aes(color = absolute_distance))+
  gghighlight(absolute_distance <= 100, use_direct_label = FALSE)+
  ggrepel::geom_label_repel(data = top5_pop_cities_near_border,
                            aes(label = paste0(city, "," , state_name), geometry = geometry),
                            stat = "sf_coordinates",
                            size = 3)+
  scale_color_viridis_c(option = "inferno", name = "Distance Difference (km)")+
  theme_map()+
  labs(title = "Most Populous City in Each State Farthest Away from the U.S. Border")

```

# Question 4
Real World Application

### 4.1 Quantifing Border Zone
```{r}
# Filter for cities in 100 mi or 160 km of border
border_zone_cities <- city_location_sp %>% 
  filter(dist_us_border_km <= 160) 

border_zone_populations <- border_zone_cities %>% 
  summarize(total_population = sum(population, na.rm = TRUE))

total_US_pop <- sum(city_location_sp$population)

percentage_pop <- border_zone_populations$total_population / total_US_pop * 100

summary_table <- data.frame(
  "Number of Cities in 100 Miles Zone" = nrow(border_zone_cities),
  "Total Population in Border Zone" = border_zone_populations$total_population,
  "Percentage of Total U.S. Population" = percentage_pop
)

summary_table

flextable(summary_table, )
```


### 4.2 Mapping Border Zone
```{r}
top10_border_zone <- border_zone_cities %>% 
  arrange(desc(population)) %>% 
  slice_head(n = 10)

ggplot()+
  geom_sf(data = city_location_sp, aes(color = dist_us_border_km))+
  gghighlight(dist_us_border_km <= 160, use_direct_label = FALSE)+
  scale_color_gradient(low = "orange", high = "darkred")+
  ggrepel::geom_label_repel(data = top10_border_zone,
                            aes(label = paste0(city, ",", state_name), geometry = geometry),
                            stat = "sf_coordinates",
                            size = 3)+
  theme_map()+
  labs(title = "10 most Populous Cities in U.S. within 100 Miles of the U.S. Border")
```

## 4.3 Instead of labeling the 10 most populous cities label the most populous cities in each state within the Danger Zone. 
```{r}
most_pop_cities_state <- border_zone_cities %>%
  filter(dist_us_border_km <= 160) %>% 
  group_by(state_name) %>% 
  slice_max(population, n =1) %>% 
  ungroup()

ggplot()+
  geom_sf(data = city_location_sp, aes(color = dist_us_border_km))+
  gghighlight(dist_us_border_km <= 160, use_direct_label = FALSE)+
  scale_color_gradient(low = "orange", high = "darkred")+
  ggrepel::geom_label_repel(data = most_pop_cities_state,
                            aes(label = paste0(city, ",", state_name), geometry = geometry),
                            stat = "sf_coordinates",
                            size = 3,
                            max.overlaps = 30)+
  theme_map()+
  labs(title = "Most Populous City in Each State within 100 Miles of the U.S. Border")
```

